---
title: "Biostat 203B Homework 5"
subtitle: Due Mar 22 @ 11:59PM
author: "Yue Wang UID:005704481"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
    eval: false
---

## Predicting ICU duration

Using the ICU cohort `mimiciv_icu_cohort.rds` you built in Homework 4, develop at least three machine learning approaches (logistic regression with enet regularization, random forest, boosting, SVM, MLP, etc) plus a model stacking approach for predicting whether a patient's ICU stay will be longer than 2 days. You should use the `los_long` variable as the outcome. You algorithms can use patient demographic information (gender, age at ICU `intime`, marital status, race), ICU admission information (first care unit), the last lab measurements before the ICU stay, and first vital measurements during ICU stay as features. You are welcome to use any feature engineering techniques you think are appropriate; but make sure to not use features that are not available at an ICU stay's `intime`. For instance, `last_careunit` cannot be used in your algorithms. 

1. Data preprocessing and feature engineering.

```{r}
library(GGally)
library(gtsummary)
library(tidyverse)
library(tidymodels)
library(ranger)
library(glmnet)
library(xgboost)
library(stacks)
library(keras)

```

```{r}
mimic_icu_cohort <- readRDS("mimic_icu_cohort.rds") %>%
  select(los_long, gender, age_at_intime, marital_status, race, 
         first_careunit, creatinine, sodium, chloride, bicarbonate,
         'white blood cell count', glucose, potassium, hematocrit, 
         'systolic non-invasive blood pressure', 
         'body temperature in Fahrenheit',
         'respiratory rate', 'diastolic non-invasive blood pressure', 
         'heart rate', subject_id, hadm_id, stay_id) %>% 
  print(width = Inf)
  
```
```{r}
mimic_icu_cohort$los_long <- as.factor(mimic_icu_cohort$los_long)

```

```{r}
mimic_icu_cohort %>% tbl_summary(by = los_long)
```

2. Partition data into 50% training set and 50% test set. Stratify partitioning according to `los_long`. For grading purpose, sort the data by `subject_id`, `hadm_id`, and `stay_id` and use the seed `203` for the initial data split. Below is the sample code.

```{r}
# sort
mimic_icu_cohort <- mimic_icu_cohort %>%
  arrange(subject_id, hadm_id, stay_id)

data_split <- initial_split(
  mimic_icu_cohort, 
  # stratify by los_long
  strata = "los_long", 
  prop = 0.5
  )

```

```{r}
mimic_other <- training(data_split)
dim(mimic_other)
```

```{r}
mimic_test <- testing(data_split)
dim(mimic_test)
```
**Answer:** There are missing values in 'marital_status', lab measurements and
vital measurements, so we will use the imputation.
```{r}
# Recipe
mimic_recipe <- 
  recipe(
    los_long ~ ., 
    data = mimic_other
  ) %>%
  # imputation 
  step_impute_mean(creatinine, sodium, chloride, bicarbonate,
         'white blood cell count', glucose, potassium, hematocrit, 
         'systolic non-invasive blood pressure', 
         'body temperature in Fahrenheit',
         'respiratory rate', 'diastolic non-invasive blood pressure', 
         'heart rate') %>%
  step_impute_mode(marital_status) %>%
  # create traditional dummy variables
  step_dummy(all_nominal_predictors()) %>%
  # zero-variance filter
  step_zv(all_numeric_predictors()) %>%
  # center and scale numeric data
  step_normalize(all_numeric_predictors()) %>%
  # estimate the means and standard deviations
  # prep(training = Heart_other, retain = TRUE) %>%
  print()

```
3. Train and tune the models using the training set.

4. Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each machine learning algorithm and the model stacking. Interpret the results. What are the most important features in predicting long ICU stays? How do the models compare in terms of performance and interpretability?

**Answer:** 1. Logistic Regression
```{r}
# Model
logit_mod <- 
  logistic_reg(
    penalty = tune(), 
    mixture = tune()
  ) %>% 
  set_engine("glmnet", standardize = FALSE) %>%
  print()
```

```{r}
# Workflow
logit_wf <- workflow() %>%
  add_recipe(mimic_recipe) %>%
  add_model(logit_mod)
logit_wf
```

```{r}
# grid
param_grid_logit <- grid_regular(
  penalty(range = c(-6, 3)), 
  mixture(),
  levels = c(100, 5)
  ) %>%
  print()

```

```{r}
# CV
set.seed(203)

folds_logit <- vfold_cv(mimic_other, v = 5)
folds_logit

```

```{r}
(logit_fit <- logit_wf %>%
  tune_grid(
    resamples = folds_logit,
    grid = param_grid_logit,
    metrics = metric_set(roc_auc, accuracy)
    )) %>%
  system.time()

```

```{r}
logit_fit
```

```{r}
logit_fit %>%
  collect_metrics() %>%
  print(width = Inf) %>%
  filter(.metric == "roc_auc") %>%
  ggplot(mapping = aes(x = penalty, y = mean, color = factor(mixture))) +
  geom_point() +
  labs(x = "Penalty", y = "CV AUC") +
  scale_x_log10()
```

```{r}
logit_fit %>%
  show_best("roc_auc")
```

```{r}
best_logit <- logit_fit %>%
  select_best("roc_auc")
best_logit
```

```{r}
# Final workflow
final_wf_logit <- logit_wf %>%
  finalize_workflow(best_logit)
final_wf_logit
```

```{r}
# Fit the whole training set, then predict the test cases
final_fit_logit <- 
  final_wf_logit %>%
  last_fit(data_split)
final_fit_logit
```

The area under ROC curve and the accuracy for logistic regression
model are shown below:
```{r}
# Test metrics
final_fit_logit %>%
  collect_metrics()
```

**Answer:** 2. Random Forest
```{r}
# Model
rf_mod <- 
  rand_forest(
    mode = "classification",
    mtry = tune(),
    trees = tune()
  ) %>%
  set_engine("ranger")
rf_mod

```

```{r}
# Workflow
rf_wf <- workflow() %>%
  add_recipe(mimic_recipe) %>%
  add_model(rf_mod)
rf_wf

```

```{r}
param_grid_rf <- grid_regular(
  trees(range = c(80L, 120L)), 
  mtry(range = c(1L, 5L)),
  levels = c(3, 3)
  )
param_grid_rf
```

```{r}
set.seed(203)

folds_rf <- vfold_cv(mimic_other, v = 5)
folds_rf

```

```{r}
rf_fit <- rf_wf %>%
  tune_grid(
    resamples = folds_rf,
    grid = param_grid_rf,
    metrics = metric_set(roc_auc, accuracy)
    )
rf_fit

```

```{r}
rf_fit %>%
  collect_metrics() %>%
  print(width = Inf) %>%
  filter(.metric == "roc_auc") %>%
  ggplot(mapping = aes(x = trees, y = mean, color = factor(mtry))) +
  geom_point() + 
  # geom_line() + 
  labs(x = "Num. of Trees", y = "CV AUC")
```

```{r}
rf_fit %>%
  show_best("roc_auc")

```

```{r}
best_rf <- rf_fit %>%
  select_best("roc_auc")
best_rf
```

```{r}
# Final workflow
final_wf_rf <- rf_wf |>
  finalize_workflow(best_rf)
final_wf_rf

```

```{r}
final_fit_rf <- 
  final_wf_rf %>%
  last_fit(data_split)
final_fit_rf
```

The area under ROC curve and the accuracy for random forest model 
are shown below:
```{r}
final_fit_rf %>% 
  collect_metrics()
```


**Answer:** 3. XGBoost
```{r}
gb_mod <- 
  boost_tree(
    mode = "classification",
    trees = 1000, 
    tree_depth = tune(),
    learn_rate = tune()
  ) %>% 
  set_engine("xgboost")
gb_mod
```

```{r}
gb_wf <- workflow() %>%
  add_recipe(mimic_recipe) %>%
  add_model(gb_mod)
gb_wf
```

```{r}
param_grid_gb <- grid_regular(
  tree_depth(range = c(1L, 3L)),
  learn_rate(range = c(-5, 2), trans = log10_trans()),
  levels = c(3, 3)
  )
param_grid_gb
```

```{r}
set.seed(203)

folds_gb <- vfold_cv(mimic_other, v = 3)
folds_gb
```

```{r}
gb_fit <- gb_wf %>%
  tune_grid(
    resamples = folds_gb,
    grid = param_grid_gb,
    metrics = metric_set(roc_auc, accuracy)
    )
gb_fit
```

```{r}
gb_fit %>%
  collect_metrics() %>%
  print(width = Inf) %>%
  filter(.metric == "roc_auc") %>%
  ggplot(mapping = aes(x = learn_rate, y = mean, color = factor(tree_depth))) +
  geom_point() +
  labs(x = "Learning Rate", y = "CV AUC") +
  scale_x_log10()
```

```{r}
gb_fit %>%
  show_best("roc_auc")
```

```{r}
best_gb <- gb_fit %>%
  select_best("roc_auc")
best_gb
```

```{r}
# Final workflow
final_wf_gb <- gb_wf %>%
  finalize_workflow(best_gb)
final_wf_gb
```

```{r}
final_fit_gb <- 
  final_wf_gb %>%
  last_fit(data_split)
final_fit_gb
```

The area under ROC curve and the accuracy for XGBoost model are shown below:
```{r}
final_fit_gb %>% 
  collect_metrics()
```

**Answer:** 4. Model Stacking
```{r}
set.seed(203)
folds_stack <- vfold_cv(mimic_other, v = 2)
```


```{r}
# logistic regression
stack_grid_logit <- grid_regular(
  penalty(range = c(0, 3)), 
  mixture(),
  levels = c(10, 5)
  )

logit_res <- 
  tune_grid(
    object = logit_wf, 
    resamples = folds_stack, 
    grid = stack_grid_logit,
    control = control_stack_grid()
  )
logit_res
```

```{r}
# random forest
stack_grid_rf <- grid_regular(
  trees(range = c(20L, 30L)), 
  mtry(range = c(1L, 2L)),
  levels = c(2, 2)
  )

rf_res <- 
  tune_grid(
    object = rf_wf, 
    resamples = folds_stack, 
    grid = param_grid_rf,
    control = control_stack_grid()
  )
rf_res
```

```{r}
# XGBoost
stack_grid_gb <- grid_regular(
  tree_depth(range = c(1L, 2L)),
  learn_rate(range = c(-1, 0), trans = log10_trans()),
  levels = c(2, 2)
  )

gb_res <- 
  tune_grid(
    object = gb_wf, 
    resamples = folds_stack, 
    grid = stack_grid_gb,
    control = control_stack_grid()
  )
gb_res
```

```{r}
mimic_model_st <- 
  stacks() %>%
  add_candidates(logit_res) %>%
  add_candidates(rf_res) %>%
  add_candidates(gb_res) %>%
  blend_predictions(
    penalty = 10^(-4),
    metrics = c("roc_auc")
    ) %>%
  # fit the candidates with nonzero stacking coefficients
  fit_members()
mimic_model_st
```

```{r}
autoplot(mimic_model_st)
```

```{r}
autoplot(mimic_model_st, type = "members")
autoplot(mimic_model_st, type = "weights")

```

```{r}
mimic_pred <- mimic_test %>%
  bind_cols(predict(mimic_model_st, ., type = "prob")) %>%
  print(width = Inf)
```

```{r}
yardstick::roc_auc(
  mimic_pred,
  truth = los_long,
  contains(".pred_TRUE")
  )
```


```{r}
mimic_pred <-
  mimic_test %>%
  select(los_long) %>%
  bind_cols(
    predict(
      mimic_model_st,
      mimic_test,
      type = "class",
      members = TRUE
      )
    ) %>%
  print(width = Inf)
```

```{r}
map(
  colnames(mimic_pred),
  ~mean(mimic_pred$los_long == pull(mimic_pred, .x))
  ) %>%
  set_names(colnames(mimic_pred)) %>%
  as_tibble() %>%
  pivot_longer(c(everything(), -los_long))
```